---
title: "Iris Dataset Analysis"
author: "Ashley King"
date: '`r format(Sys.time(), "%b %d, %Y at %X")`'
output:
  bookdown::html_document2:
    highlight: textmate
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)

library(tidyverse)
library(resampledata)
library(knitr)
library(gridExtra)
library(ggplot2)
library(dplyr)
library(gapminder)
library(skimr)
library(ggcorrplot)
library(caret)
library(rpart)
library(rpart.plot)
library(pROC)
```

# Iris dataset analysis

The iris dataset contains measurements about iris flowers, with the aim of classifying the species.
We will split the data into a train and a test set, and then do an EDA on the train set. In the end, we will create a decision tree model to predict the train set, and test the validity on the test set.

# Splitting the data into train and test

```{r}
set.seed(1)
data(iris)
train <- createDataPartition(iris$Species, p = .8, list = FALSE, times = 1)
trainIrisData <- iris[train,]
testIrisData <- iris[-train,]
```

# Looking at the raw data

```{r}
trainIrisData %>% 
  skim(-Species)
```

+ We have 4 variables, Sepal.Length, Sepal.Width, Petal.Length and Petal.Width
+ We do not have any missing variables
+ Sepal.Width has an approximate normal distribution, but we will explore the variables more visually

# Exploring our data visually

```{r}
plot(trainIrisData)
```

+ This visual explores the distribution of all variables with a scatterplot
+ Petal.Length and Petal.Width seem to have a positive linear relationship
+ Species appears to be a factor with three levels, and you can see a clear pattern within the species
    + (Looking at the right hand graphs)


## Comparing sepal length vs width by species

```{r}
ggplot(data = trainIrisData, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() + 
  facet_grid(~ Species)
```

+ Here we can see that the Setosa flower on average has the highest width and the smallest length.
+ The Versicolor has a smaller width and length on average than the Virginica.

## Comparing Petal length vs width by species

```{r}
ggplot(data = trainIrisData, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point() + 
  facet_grid(~ Species)
```

+ This graph shows the large difference between petal measurements among all three Species. 
+ The Setosa, on average, has the smallest petal length and width
+ The Viriginca, on average, has the largest petal length and width

## Exploring boxplots for all variables

### Sepal Length

```{r}
ggplot(trainIrisData, aes(Species, Sepal.Length, fill=Species)) + 
        geom_boxplot()
```

+ Setosa, on average, has the shortest Sepal Length, around 5 and two high outliers
+ Versicolor, on average, has a Sepal length of around 6
+ Virginica has one low outlier around 5, while the average Sepal Length is around 6.5

### Sepal Width 

```{r}
ggplot(trainIrisData, aes(Species, Sepal.Width, fill=Species)) + 
        geom_boxplot()
```

+ Setosa has one low outlier around 2.25 and the average Sepal Width is around 3.4
+ Versicolor has no outliers and a lower average of around 2.75
+ Virginica has one low outlier around 2.25 and two high outliers above 3.6, with an average Sepal Width of 3.0

### Petal Length
```{r}
ggplot(trainIrisData, aes(Species, Petal.Length, fill=Species)) + 
        geom_boxplot()
```

+ The Petal Length has a very small distribution compared to the other variables
+ The Setosa has one low outler around 1, with an average Petal Length of around 1.75
+ The Versicolor has one low outlier around 3, and an average Petal Length of around 4.25
+ The Virginica does not have any outliers, with an average Petal Length of around 5.5

### Petal Width
```{r}
ggplot(trainIrisData, aes(Species, Petal.Width, fill=Species)) + 
        geom_boxplot()
```

+ The Setosa has the smallest, on average, Petal Width, with an average Width of around 0.25. There are two high outliers around 0.5
+ Versicolor has no outliers, and an average Petal Width of around 1.25
+ The Virginica has the highest Petal Width of (on average) around 2.0

# Building our model

## Decision Tree

A decision tree works by starting at a designated root node, and making a decision based on the data. The model will either choose left or right based on if a condition is met or not. The decisions can be based on categorical data or numerical data.

```{r}
model <- rpart(Species~., data = trainIrisData, method = 'class')
rpart.plot(model)
```

The first decision that the data makes is based on the question: "Is the petal length less than 2.6?" If it is, it classifies the flower as a Setosa. If it is not, it goes to another decision node, where it asks "Is the petal width less than 1.8?" If it is, it is classified as a Versicolor, if not, it is classified as a Virginica.

### Predicting the test data

```{r}
predictTest <- predict(model, testIrisData, type = 'class')
tableTest <- table(testIrisData$Species, predictTest)
tableTest
```

+ This prediction feeds the testing data into the decision tree pictured above
+ The correct predictions are seen on the diagonal from the upper left to lower right
+ The incorrect predictions are seen on all spots that are not on the diagonal, with 2 Species incorrectly predicted.

## Accuracy

```{r}
mean(predictTest == testIrisData$Species)
```

+ We have a 93% accuraccy

## Error
```{r}
mean(predictTest != testIrisData$Species)
```

+ We only have a 3.33% error rate

## Confusion Matrix

```{r}
confusionMatrix(predictTest, testIrisData$Species)
```

 + The confusion matrix summarizes the two points we had before
 + We have an Accuracy of 96.67%
 + The Kappa is 95%, which in short is the accuracy accounted for chance
 
 
 